**Bayesian statistics for longitudinal studies in biomedical research**

### 2020-11-03 Manuscripts/Manuscript_AM_v1.Rmd
updated
Updated manuscript, need to work on first section

bibliography: refs.bib
output:
  word_document:
    reference_docx: docx_template.docx
  pdf_document: 
    fig_caption: yes
  css: style.css
  html_document: default
mainfont: Times New Roman
link-citations: yes
subtitle: '_Their application and use in biomedical research_'
biblio-style: apalike
---

@@

Recently, _linear mixed effects models_ (LMEMs) have been used by certain groups to analyze biomedical longitudinal data [@vishwanath2009;@skala2010]. Briefly, these models incorporate _fixed_ effects, which correspond to the levels of experimental factors in the study (e.g. the different drug regimens in a clinical trial), and _random_ effects, which account for random variation within the population [@pinheiro2006]. These models are more flexible than rm-ANOVA as they can accomodate missing observations in the data, and  allow to model the covariance of the parameters in different manners[@west2014]. On the other hand, they impose restrictions in the distribution of the errors and of the random effects, and assume linearity between the response and the parameters[@schielzeth2020;@pinheiro2006].

This assumption of linearity in LMEMs is also present in rm-ANOVA, and therefore both models restrict the inferences they can extract from a longitudinal study when the data does not follow a linear trend.In both cases the model does not provide a consistent fit with the trend of the data. In such circumstances, it is even possible to obtain a "significant" _p-value_( _p_<0.05) due to the well-known variability of this metric [@nuzzo2014;@halsey2019], but the model will lack predictive power and this in turn will compromise the extent of the inferences that can be derived. The non-linear behavior in longitudinal data is exemplified in different biomedical studies, including response to radio/chemotherapy in preclinical and clinical settings  [@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994].  Since both the _frequentist_ (rm-ANOVA) and LMEM approaches are limited in the analysis of non-linear longitudinal information, there is a need to use statistical tools that allow both a consistent fit with the data and that allow adequate inferences.

In this regard, _Bayesian statistics_ represent a relatively new field in statistics that does not rely on _p-values_ and hypothesis tests to analyze information. Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine the outcome in regard to significance and are able to expand the comparisons and inferences derived form the analysis. On the other hand, the shift that Bayesian theory represents from the traditional _frequentist_ statistical view in research and the computational tools required for the implementation of this type of models have limited their use in the biomedical research community. However, the increasing recognition of the advantages of Bayesian statistics and the current development in computational tools are enabling researchers to transition to the field more rapidly.

@@

>>>>

bibliography: refs.bib
output:
  html_document: default
  pdf_document: 
    fig_caption: yes
  css: style.css
  word_document:
    reference_docx: docx_template.docx
mainfont: Times New Roman
link-citations: yes
subtitle: '_Their application and use in biomedical research_'
biblio-style: apalike
---

@@

Recently, _linear mixed effects models_ (LMEMs) have been used by certain groups to analyze biomedical longitudinal data [@vishwanath2009;@skala2010]. Briefly, these models incorporate _fixed_ effects, which correspond to the levels of experimental factors in the study (e.g. the different drug regimens in a clinical trial), and _random_ effects, which account for random variation within the population [@pinheiro2006]. These models are more flexible than rm-ANOVA as they can accommodate missing observations in the data, and  allow different modeling strategies for the covariance[@west2014]. On the other hand, they impose restrictions in the distribution of the errors and of the random effects, and assume linearity between the response and the parameters[@schielzeth2020;@pinheiro2006].

The assumption of linearity is a key component of LMEMs and rm-ANOVA, and therefore both models restrict the inferences they can extract from longitudinal data when it does not follow a linear trend. In biomedical research, such behavior in longitudinal data does arise and has been exemplified in different studies, including response to radio/chemotherapy in preclinical and clinical settings  [@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994]. In such situations, an estimated model via rm-ANOVA or LMEMs may not provide a consistent fit with the observed variations in the variables of interest. Moreover, although it is possible to obtain a "significant" _p-value_( _p_<0.05) due to the well-known variability of this metric [@nuzzo2014;@halsey2019], the model will lack predictive power and this in turn will compromise the extent of the inferences that can be derived from it.

Since both the _frequentist_ (rm-ANOVA) and LMEM approaches are limited in the analysis of non-linear longitudinal information, there is a need for biomedical researchers to use statistical tools that allow the data to determine the fit of the model while enabling  inferences that are  both adequate and consistent from a statistical standpoint.

Generalized additive models (GAMs) are a subset of generalized linear models that use _smooth functions_ (hencefort _smooths_) to estimate the parameters of a model. They have been used in palaeolimnology, ecology and clinical studies to model longitudinal data[@woolway2016;@hefley2017;@ko2007]. Briefly, GAMs use a combination of multiple functions (basis functions)to construct the smooths of the model [@wood2017]. Their main advantage over LMEMs and rm-ANOVA is that the model specification is directed by the _smooths_ rather than by a parametric relationship. This allows a consistent fit of the model with the data, and estimations of significance using the terms of the model. However, certain assumptions about the data are necessary: a normal distribution and constant variance of the residuals with the mean response. Therefore, GAMs provide a more suitable statistical method to analyze biomedical longitudinal data, when these assumptions are met.

If the assumptions of GAMs do not hold,  the field of _Bayesian statistics_ represents a relatively new area of Statistics that does not rely on _p-values_ and hypothesis tests to analyze information. Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine the outcome in regard to significance and are able to expand the number of comparisons and inferences derived form the analysis. On the other hand, the shift that Bayesian statistics represent from the traditional _frequentist_ statistical view in research, the computational tools required for its implementation, and the underlying mathematical theory have limited the use of this approach in the biomedical research community. However,Bayesian theory is intuitive and shares some principles with _frequentist_ statistics, and there is an increasing use and recognition of the advantages of their use across different areas of biomedical research such as clinical trial design and imaging[@biswas2009;@kelter2020;@kwon2020;@zhou2017]. Additionaly,the current development in computational tools, specifically the programming language $\textsf{R}$, enable a rapid implementation of Bayesian models.

@@

https://www.frontiersin.org/articles/10.3389/fevo.2018.00149/full
